Implicit Neural Representation for Potential Field Geophysics

### 2023-10-10 discussion.tex
Add short headings to suit ToC

\section{SR impact on interpretation}
The super-resolution outputs from the methods presented in this thesis have improved structural accuracy over the low-resolution grids.
This is likely to have a positive impact on the detection of features such as lineaments and unit boundaries for geological interpretation from magnetic data.
However, despite the reduction of introduced structural inaccuracies with the use of open access training data in \Cref{ch:paper2}, it is likely that some features with specific structural geometries will still be incorrectly super-resolved.
A domain expert may be able to recognise structures that are inconsistent using auxiliary information and the regional geological context, but machine processed super-resolution grids will use these data as provided.
In order to improve the adoption of SR for geological interpretation from potential field data, a rigorous study of the impact of SR on interpretation outcomes is required.
This could be a geological mapping task, such as the reinterpretation of regional geology using super-resolution upscaled regional magnetic TMI grids.
Comparison could then be made for the accuracy and benefit of a super-resolution method compared to the existing low-resolution interpretation.
However, such analysis will be difficult to perform with respect to different interpreters' variability in pattern recognition, especially in the context of their confidence in the reliability of the super-resolution product.
Previously, \textcite{sivarajahIdentifyingEffectiveInterpretation2013} reported significant variability between individuals when detecting geological targets from magnetic geophysics data.
A more reliable, repeatable and quantifiable approach would use automated pattern recognition methods developed for geophysical features.
For example, the automated lineament detection technique of \textcite{holdenAutomaticIdentificationResponses2011,holdenAutomatedAnalysisRegional2008} could be applied to low-resolution and SR grids, to compare their delineated features.

\section{SR model construction}
Several models were implemented throughout \Cref{ch:paper1,ch:paper2}, namely RDN, ESRGAN+, and LTE\@.
% However, the most important choice in developing a performant super-resolution neural network for real-world aeromagnetic survey data was the choice of training data.
\Cref{ch:paper1} was prepared prior the release of a large and varied synthetic geology training dataset \parencite{jessellNoddyverseMassiveData2022}.
It was seen in \Cref{ch:paper2} that this data improved model performance for a range of features by using transfer learning for pre-training.
% This transfer learning approach for training data is also found to be useful for RDN and ESRGAN\@.
Here, the use of synthetic and real-world line spacing sampled data approach used with LTE in \Cref{ch:paper2} are adopted for the RDN+ and ESRGAN+ models.
\Cref{fig:modcomp} shows RDN and ESRGAN+ in predicting accurate features when upscaling \qty{320}{\m} data to \qty{80}{\m} line spacing.
It can be inferred that the challenge of super-resolution for these tasks is predominantly a question of providing suitable and sufficient training data, and is less constrained to neural network choices.

>>>>

\section[Super-resolution impact on interpretation]{SR impact on interpretation}
The super-resolution outputs from the methods presented in this thesis have improved structural accuracy over the low-resolution grids.
This is likely to have a positive impact on the detection of features such as lineaments and unit boundaries for geological interpretation from magnetic data.
However, despite the reduction of introduced structural inaccuracies with the use of open access training data in \Cref{ch:paper2}, it is likely that some features with specific structural geometries will still be incorrectly super-resolved.
A domain expert may be able to recognise structures that are inconsistent using auxiliary information and the regional geological context, but machine processed super-resolution grids will use these data as provided.
In order to improve the adoption of SR for geological interpretation from potential field data, a rigorous study of the impact of SR on interpretation outcomes is required.
This could be a geological mapping task, such as the reinterpretation of regional geology using super-resolution upscaled regional magnetic TMI grids.
Comparison could then be made for the accuracy and benefit of a super-resolution method compared to the existing low-resolution interpretation.
However, such analysis will be difficult to perform with respect to different interpreters' variability in pattern recognition, especially in the context of their confidence in the reliability of the super-resolution product.
Previously, \textcite{sivarajahIdentifyingEffectiveInterpretation2013} reported significant variability between individuals when detecting geological targets from magnetic geophysics data.
A more reliable, repeatable and quantifiable approach would use automated pattern recognition methods developed for geophysical features.
For example, the automated lineament detection technique of \textcite{holdenAutomaticIdentificationResponses2011,holdenAutomatedAnalysisRegional2008} could be applied to low-resolution and SR grids, to compare their delineated features.

\section[Super-resolution model construction]{SR model construction}
Several models were implemented throughout \Cref{ch:paper1,ch:paper2}, namely RDN, ESRGAN+, and LTE\@.
% However, the most important choice in developing a performant super-resolution neural network for real-world aeromagnetic survey data was the choice of training data.
\Cref{ch:paper1} was prepared prior the release of a large and varied synthetic geology training dataset \parencite{jessellNoddyverseMassiveData2022}.
It was seen in \Cref{ch:paper2} that this data improved model performance for a range of features by using transfer learning for pre-training.
% This transfer learning approach for training data is also found to be useful for RDN and ESRGAN\@.
Here, the use of synthetic and real-world line spacing sampled data approach used with LTE in \Cref{ch:paper2} are adopted for the RDN+ and ESRGAN+ models.
\Cref{fig:modcomp} shows RDN and ESRGAN+ in predicting accurate features when upscaling \qty{320}{\m} data to \qty{80}{\m} line spacing.
It can be inferred that the challenge of super-resolution for these tasks is predominantly a question of providing suitable and sufficient training data, and is less constrained to neural network choices.

### 2023-10-10 intro.tex
Add short headings to suit ToC

\section{Traditional methods for enhancing resolution}
As outlined in \Cref{sec:introgrids}, resolution is a product of sample distance and density, and gridded rasters have a specific number of cells in each axis depending on the spatial extent.
Upsampling raster data to larger height and width dimensions is commonplace, and generally performed by numerical image filters such as the cubic spline \parencite{keysCubicConvolutionInterpolation1981} or Lanczos filter \parencite{lanczos1988applied}.
Some of these interpolation methods are the basis of the gridding methods previously outlined, and simply selecting a smaller cell size during gridding is sufficient to achieve a larger image raster.
These filters are adept at imputing the required cell values while preserving the frequency content of the original raster.
However, no attempt is made at accurately increasing the high-frequency components of the raster, which are required by our definition for high-resolution geophysics.
Without sampling at a smaller separation or closer line spacing, it is still possible to increase the high-frequency content of aeromagnetic grids.

@@

\section{Geophysical data in machine learning}
\label{sec:introdata}
% e.g. limited training data to deal with diverse geological features, geology various regionally - different potential fields features and range, etc....
Natural image raster data are ubiquitous in computer vision machine learning.
These are most commonly three channel arrays containing normalised brightness values in the red, green and blue channels.
Resolution in these rasters is typically a product of an imaging system or software defined grid with regular sampling.
Like in geophysics, these raster data have an array height and width, which is referred to as resolution.
However, the lack of spatial quantification for the size of natural images confuses the term with its definition for geophysical resolution.

@@

\section{Machine learning in geoscience}
\label{sec:introgeoml}

>>>>

\subsection{Traditional methods for enhancing resolution}
As outlined in \Cref{sec:introgrids}, resolution is a product of sample distance and density, and gridded rasters have a specific number of cells in each axis depending on the spatial extent.
Upsampling raster data to larger height and width dimensions is commonplace, and generally performed by numerical image filters such as the cubic spline \parencite{keysCubicConvolutionInterpolation1981} or Lanczos filter \parencite{lanczos1988applied}.
Some of these interpolation methods are the basis of the gridding methods previously outlined, and simply selecting a smaller cell size during gridding is sufficient to achieve a larger image raster.
These filters are adept at imputing the required cell values while preserving the frequency content of the original raster.
However, no attempt is made at accurately increasing the high-frequency components of the raster, which are required by our definition for high-resolution geophysics.
Without sampling at a smaller separation or closer line spacing, it is still possible to increase the high-frequency content of aeromagnetic grids.

@@

\section{Geophysics and machine learning}
\label{sec:introdata}
% e.g. limited training data to deal with diverse geological features, geology various regionally - different potential fields features and range, etc....
Natural image raster data are ubiquitous in computer vision machine learning.
These are most commonly three channel arrays containing normalised brightness values in the red, green and blue channels.
Resolution in these rasters is typically a product of an imaging system or software defined grid with regular sampling.
Like in geophysics, these raster data have an array height and width, which is referred to as resolution.
However, the lack of spatial quantification for the size of natural images confuses the term with its definition for geophysical resolution.

@@

\subsection{Machine learning in geoscience}
\label{sec:introgeoml}

### 2023-10-10 paper2.tex
Add short headings to suit ToC

\subsection{Quantitative Measure}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}

\section{Results}
\label{sec:2results}
Model performance is quantified using FSIMi in Cref{tab:2metrics}, and it is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids over the baseline model.

@@

\section{Conclusion}
\label{sec:2conclusion}
An approach for upscaling geologically diverse geophysical survey grid data is proposed, which adopts transfer learning from synthetic data to better generalise super-resolution across different geological domains.
Using the preeminent LTE neural network, synthetic magnetic forward models were used to prepare a baseline model capable of upscaling survey grids at four times scale.
Further, this baseline model was fine-tuned with open access survey grid data, prepared with a low-resolution transform appropriate for aeromagnetic geophysics.
Low-resolution features incorrectly enhanced by the synthetic baseline model were correctly resolved following fine-tune training with these data, and the model is capable of upscaling grids with \qty{320}{\m} line spacing to \qty{80}{\m} line spacing with an average FSIM score of \num{0.9731}.

>>>>

\section{Results}
\label{sec:2results}
\subsubsection{Quantitative Measure}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}
Model performance is quantified using FSIMi in Cref{tab:2metrics}, and it is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids over the baseline model.

@@

\section{Conclusions}
\label{sec:2conclusion}
An approach for upscaling geologically diverse geophysical survey grid data is proposed, which adopts transfer learning from synthetic data to better generalise super-resolution across different geological domains.
Using the preeminent LTE neural network, synthetic magnetic forward models were used to prepare a baseline model capable of upscaling survey grids at four times scale.
Further, this baseline model was fine-tuned with open access survey grid data, prepared with a low-resolution transform appropriate for aeromagnetic geophysics.
Low-resolution features incorrectly enhanced by the synthetic baseline model were correctly resolved following fine-tune training with these data, and the model is capable of upscaling grids with \qty{320}{\m} line spacing to \qty{80}{\m} line spacing with an average FSIM score of \num{0.9731}.

### 2023-10-10 paper3.tex
Add short headings to suit ToC

\section{Automatic Differentiation}
Automatic differentiation (AD) refers to one of several methods for computing derivatives, specifically to the method of accumulating derivatives during function evaluation \parencite{baydinAutomaticDifferentiationMachine2018}.
This method underpins modern machine learning frameworks such as Pytorch \parencite{paszkePyTorchImperativeStyle2019}, which is used to construct the network architecture and perform training in this work.
In these frameworks, training is performed by evaluating a set of inputs using the current model parameters during a forward pass, while accumulating the partial derivatives of the result with respect to those parameters, to optimise an objective (loss) function.
These gradients inform parameter update operations to minimise the model loss.
However, automatic differentiation can also evaluate the derivatives of the network output with respect to the inputs, using the same high-performance AD framework.
This can directly provide the partial derivatives of the potential field with respect to its coordinates \(\frac{\partial\phi{}}{\partial x}, \frac{\partial \phi{}}{\partial y}, \frac{\partial \phi{}}{\partial z}\), as well as higher order derivatives, without first requiring gridding.

>>>>

\subsection{Automatic Differentiation}
Automatic differentiation (AD) refers to one of several methods for computing derivatives, specifically to the method of accumulating derivatives during function evaluation \parencite{baydinAutomaticDifferentiationMachine2018}.
This method underpins modern machine learning frameworks such as Pytorch \parencite{paszkePyTorchImperativeStyle2019}, which is used to construct the network architecture and perform training in this work.
In these frameworks, training is performed by evaluating a set of inputs using the current model parameters during a forward pass, while accumulating the partial derivatives of the result with respect to those parameters, to optimise an objective (loss) function.
These gradients inform parameter update operations to minimise the model loss.
However, automatic differentiation can also evaluate the derivatives of the network output with respect to the inputs, using the same high-performance AD framework.
This can directly provide the partial derivatives of the potential field with respect to its coordinates \(\frac{\partial\phi{}}{\partial x}, \frac{\partial \phi{}}{\partial y}, \frac{\partial \phi{}}{\partial z}\), as well as higher order derivatives, without first requiring gridding.
